Hyperbolic SMC: a Variational Proposal for Sequential Monte Carlo Phylogenetic Inference

### 2024-07-31 paper/paper.tex
Remove comment
Add two-phased training section

% additional packages
\usepackage{siunitx}
\usepackage{physics}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

% TODO replace this with Hyperbolic SMC
\begin{algorithm}
  \caption{Hyperbolic Sequential Monte Carlo}
  \label{alg:hyperbolic-smc}
  \begin{algorithmic}[1]
    \State \textbf{Input:} $\theta = (Q, v_0^{1:N}, \sigma)$
    \State Initialize $\forall k \in [K], s_0^k \leftarrow \bot, w_0^k \leftarrow 1/K$
    \State Initialize $\forall k \in [K], \forall n \in [N], v_0^{k,n} \leftarrow v_0^n$
    \For{$r = 1$ \textbf{to} $R = N - 1$}
    \For{$k \in [K]$}
    \State Resample particle
    \[
      \mathbb{P}(i_\text{resample} = i) = \frac{w_{r-1}^i}{\sum_{l=1}^K w_{r-1}^l} \; , \forall \, i \in [K]
    \]
    \State Sample two trees uniformly
    \[
      \mathbb{P}(t_\text{left} = i, t_\text{right} = j) = \frac{1}{\binom{N-r+1}{2}} , \; \forall \, i < j
    \]
    \State Form internal embedding $v_\text{parent}$ from $v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}$
    \State Compute parent-child distances
    \[
      d_\text{left} = d\qty(v_{r-1}^{k,t_\text{left}}, v_\text{parent}) \;,\; d_\text{right} = d\qty(v_{r-1}^{k,t_\text{right}}, v_\text{parent})
    \]
    \State Sample branch lengths from log-normal distributions (with reparameterization trick)
    \[
      b_\text{left} = \exp(\log(d_\text{left}) + \sigma Z_\text{left}) \;,\; b_\text{right} = \exp(\log(d_\text{right}) + \sigma Z_\text{right})
    \]
    \[
      Z_\text{left}, Z_\text{right} \sim \mathcal{N}(0, 1)
    \]
    \State Form new partial state $s_r^k$ by merging trees $t_\text{left}, t_\text{right}$
    \State Compute proposal probability
    \[
      q(s_r^k | s_{r-1}^{i_\text{resample}}) = \mathbb{P}(t_\text{left} = i, t_\text{right} = j) \cdot f_\text{log-normal}(b_\text{left} | d_\text{left}, \sigma) \cdot f_\text{log-normal}(b_\text{right} | d_\text{right}, \sigma)
    \]
    \State Compute weight
    \[
      w_r^k = \frac{\pi(s_r^k) \cdot \nu^-(s_{r-1}^{i_\text{resample}})}{\pi(s_{r-1}^{i_\text{resample}}) \cdot  q(s_r^k | s_{r-1}^{i_\text{resample}})}
    \]
    \State
    Update tree root embeddings
    \[
      \qty{v_r^{k,1}, \cdots, v_r^{k,N-r}} \leftarrow \qty(\qty{v_{r-1}^{k,1}, \cdots, v_{r-1}^{k,N-r+1}} \setminus \qty{v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}}) \cup \qty{v_\text{parent}}
    \]
    \EndFor
    \EndFor
    \State \Return $s_{1:R}^{1:K}, w_{1:R}^{1:K}$
  \end{algorithmic}
\end{algorithm}

@@

>>>>

% additional packages
\usepackage{siunitx}
\usepackage{physics}
\usepackage{cancel}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\begin{algorithm}
  \caption{Hyperbolic Sequential Monte Carlo}
  \label{alg:hyperbolic-smc}
  \begin{algorithmic}[1]
    \State \textbf{Input:} $\theta = (Q, v_0^{1:N}, \sigma)$
    \State Initialize $\forall k \in [K], s_0^k \leftarrow \bot, w_0^k \leftarrow 1/K$
    \State Initialize $\forall k \in [K], \forall n \in [N], v_0^{k,n} \leftarrow v_0^n$
    \For{$r = 1$ \textbf{to} $R = N - 1$}
    \For{$k \in [K]$}
    \State Resample particle
    \[
      \mathbb{P}(i_\text{resample} = i) = \frac{w_{r-1}^i}{\sum_{l=1}^K w_{r-1}^l} \; , \forall \, i \in [K]
    \]
    \State Sample two trees uniformly
    \[
      \mathbb{P}(t_\text{left} = i, t_\text{right} = j) = \frac{1}{\binom{N-r+1}{2}} , \; \forall \, i < j
    \]
    \State Form internal embedding $v_\text{parent}$ from $v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}$
    \State Compute parent-child distances
    \[
      d_\text{left} = d\qty(v_{r-1}^{k,t_\text{left}}, v_\text{parent}) \;,\; d_\text{right} = d\qty(v_{r-1}^{k,t_\text{right}}, v_\text{parent})
    \]
    \State Sample branch lengths from log-normal distributions (with reparameterization trick)
    \[
      b_\text{left} = \exp(\log(d_\text{left}) + \sigma Z_\text{left}) \;,\; b_\text{right} = \exp(\log(d_\text{right}) + \sigma Z_\text{right})
    \]
    \[
      Z_\text{left}, Z_\text{right} \sim \mathcal{N}(0, 1)
    \]
    \State Form new partial state $s_r^k$ by merging trees $t_\text{left}, t_\text{right}$
    \State Compute proposal probability
    \[
      q\qty(s_r^k | s_{r-1}^{i_\text{resample}}) = \mathbb{P}(t_\text{left} = i, t_\text{right} = j) \cdot f_\text{log-normal}(b_\text{left} | d_\text{left}, \sigma) \cdot f_\text{log-normal}(b_\text{right} | d_\text{right}, \sigma)
    \]
    \State Compute weight
    \[
      w_r^k = \frac{\pi\qty(s_r^k) \cdot \nu^-\qty(s_{r-1}^{i_\text{resample}})}{\pi\qty(s_{r-1}^{i_\text{resample}}) \cdot  q\qty(s_r^k | s_{r-1}^{i_\text{resample}})}
    \]
    \State
    Update tree root embeddings
    \[
      \qty{v_r^{k,1}, \cdots, v_r^{k,N-r}} \leftarrow \qty(\qty{v_{r-1}^{k,1}, \cdots, v_{r-1}^{k,N-r+1}} \setminus \qty{v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}}) \cup \qty{v_\text{parent}}
    \]
    \EndFor
    \EndFor
    \State \Return $s_{1:R}^{1:K}, w_{1:R}^{1:K}$
  \end{algorithmic}
\end{algorithm}

@@

Hyperbolic SMC converges especially fast when the variance in the likelihood estimate is minimized. A large share of the variance is due to branch length sampling. One can imagine an alternate Hyperbolic SMC proposal where parent-child embedding distance are taken directly as branch lengths, with no sampling. We must drop the continuous $f_\text{log-normal}(b_\text{left/right} | d_\text{left/right}, \sigma)$ terms from $q(s_r^k | s_{r-1}^{i_\text{resample}})$, but this introduces a bias in the likelihood estimate. To see this effect more clearly, consider the expectation (over topology $\tau$ and branch lengths $\mathcal B$) of the log weights:

\begin{equation}
  \mathbb{E}_{\tau, \mathcal B \sim q\qty(\cdot | s_{r-1}^{i_\text{resample}})} \qty[\log w_r^k]
  = \dots - \mathbb{E}_{\tau, \mathcal B} \qty[\log q\qty(\tau | s_{r-1}^{i_\text{resample}})]  - \cancel{\mathbb{E}_{\tau, \mathcal B} \qty[\log q\qty(\mathcal B | \tau, s_{r-1}^{i_\text{resample}})]} \; .
\end{equation}

In the ``deterministic branch length'' setting, we drop the last term the penalizes the ELBO when the branch length distribution is sharply peaked. Namely, we overestimate the ELBO by $O(\log (1/\sigma))$, where $\sigma$ is is the interval over which sampling from $\text{Uniform}(b-\sigma, b+\sigma)$ approximates taking $b$ as the branch length.

In practice, we can form a biased likelihood estimator that---despite not being a valid ELBO---allows us to efficiently optimize the tip embeddings. In a two-phased training scheme, we first optimize embeddings while simultaneously exploring topology space under the low-variance, biased likelihood estimator. Empirically, the learned embeddings form a solid baseline for sampling under the true ELBO (i.e., with proper branch length distributions).
